Draft 1

### 2023-11-02 Lahman Analysis/Formal Analysis/Draft 1.qmd
Continued writing formal analysis

@@

Finally, it does not take an advanced degree to understand that age is an important factor in athletic ability. To an extent in MLB, seniority can act both as a blessing and a curse. Older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30. Yet, these players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing. Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in

@@

library(Lahman)
library(tidyverse)
```

```{r}
Runs.By.Year <- Lahman::Teams %>% 
  group_by(yearID) %>% 
  summarize(R = sum(R))
```

@@

ggplot(data = Runs.By.Year)+
  aes(x=yearID, y=R)+
  geom_line()+
  geom_rect(inherit.aes = FALSE, aes(xmin=1900, xmax=1920, ymin=0, ymax=26000), color="transparent", fill="red", alpha=0.005)+
  geom_rect(inherit.aes = FALSE, aes(xmin=1994, xmax=2004, ymin=0, ymax=26000), color="transparent", fill="orange", alpha=0.005)+
  scale_y_continuous(limits =c(0,26000), breaks = seq(from=0, to=25000, by=5000))+
  scale_x_continuous(breaks = seq(from=1870, to=2020, by=10))+
  labs(title = "Runs Scored in MLB by Year",
       x = "Year",
       y = "Runs Scored")
```

>>>>

library(tidyverse)
library(rstan)
library(Lahman)
library(bayesplot)
```

```{r}
#Set Stan to read the amount of cores that your computer can handel
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

```{r}
#| message: false

#Creates a DF called "Batting" which has players stats from 1973-2019

#Load the DF's from Lahman directly
Batting <- Lahman::Batting
Teams <- Lahman::Teams
People <- Lahman::People

Batting <- Batting %>% 
  #Minimum year = DH was introduced; maximum year = before COVID
  filter(yearID >= 1973 & yearID <= 2019)

#Filter teams for years considered and select only necessary columns
Teams <- Teams %>% 
  filter(yearID>=1973 & yearID <= 2019) %>% 
  select(yearID, teamID, park)

#Combine current "Batting" and "Teams" DF to associate player with stadium played at
Batting <- Batting %>% 
  right_join(Teams, by = join_by(yearID, teamID))

#Add player's age to the "Batting" DF
Batting <- Batting %>% 
  right_join(Lahman::People, by = join_by(playerID)) %>% 
  mutate(Age = yearID - birthYear, .after=yearID) %>%
  #Don't consider the people who don't have a birthday in the database
  drop_na(Age)

#Make the lgID variable into an index variable
Batting <- Batting %>% 
  mutate(lg = case_when(lgID == "NL" ~ 1,
                        lgID == "AL" ~ 2), .after=lgID)

#Make an index for the parks to add them to the model

#The index goes from 1-88. No idea how it came up with them but I kept the park name in Batting so anyone should be able to reference to see how the parks impact player performance

Batting <- Batting %>% 
  mutate(park = as.factor(park)) %>% 
  mutate(park_index = as.integer(park), .after = park)
```

@@

Finally, it does not take an advanced degree to understand that age is an important factor in athletic ability. To an extent in MLB, seniority can act both as a blessing and a curse. On the one hand, older players are less able to hit home runs because their body is starting to lose its peak athletic performance it had when it was, say, 30. On the other hand, these older players also have more experience than younger players and thus may be better able to hit home runs from strategies they have found during their time playing. Further we can visualize this by looking at the median home runs hit by each player from 1871-2022, which is displayed in @fig-HR.By.Age.All. However, this graph does not depict some of the nuances of the data well. For example, we have almost no data on how often 18 -year-olds hit HR's because they play in MLB infrequently. As can be seen with the standard error bars on the ages above 40, older players suffer a similar problem of small sample size. Instead, consider @fig-HR.By.Age.20to40 which shows only considers players who are 20-40 years old during the season. This graph clearly shows the complex non-linear relationship between player performance and age. It shows that players are generally improving in their ability to hit HR's from 20-30 years old. Then, however, we see a slight downwards trend followed by an upward trend with increasing error bars. This shows that as players age past 30 their abilities start to decline. However, the jump afterwards can probably be attributed to players who are very good, continuing their careers. If a player is very good then they are going to be able to perform at above average when they are, say, 35. However, if the player is not generally considered good, then a team will likely opt to try out a new young rookie who may not be as good in the moment but is more projectable. Thus, we see less players playing past 35 which explains the higher standard errors, but they are of a higher caliber than any other age group which is why they are still playing, thus we see their proportion as higher.

@@

```{r}
#Make the dataframe needed for Runs Scored by year graph
Runs.By.Year <- Lahman::Teams %>% 
  group_by(yearID) %>% 
  summarize(R = sum(R))
```

@@

ggplot(data = Runs.By.Year)+
  aes(x=yearID, y=R)+
  geom_line()+
  geom_rect(inherit.aes = FALSE, aes(xmin=1900, xmax=1920, ymin=0, ymax=26000), color="transparent", fill="red", alpha=0.005)+
  geom_rect(inherit.aes = FALSE, aes(xmin=1994, xmax=2004, ymin=0, ymax=26000), color="transparent", fill="orange", alpha=0.005)+
  scale_y_continuous(limits =c(0,26000), breaks = seq(from=0, to=25000, by=5000))+
  scale_x_continuous(breaks = seq(from=1870, to=2020, by=10))+
  labs(title = "Cumulative Runs Scored in MLB by Year",
       x = "Year",
       y = "Runs Scored")
```

```{r}
#Make DF which describes the HR by age across the timeline considered
HR.Prop.Age <- Batting %>% 
  group_by(Age) %>% 
  summarize(HR = sum(HR), AB = sum(AB)) %>% 
  mutate(prop = HR / AB, prop.se = sqrt(prop*(1-prop)/AB))
```

```{r}
#| label: fig-HR.By.Age.All

ggplot(data = HR.Prop.Age)+
  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - prop.se,
                                ymax = prop + prop.se))+
  labs(title = "Proportion of Home Runs per At-Bat by Age",
       x = "Age",
       y = "Proportion of HR per AB")
```

```{r}
#| label: fig-HR.By.Age.20to40

ggplot(data = HR.Prop.Age %>% filter(Age >=20 & Age <= 40))+
  geom_pointrange(mapping = aes(x = Age, y = prop, ymin = prop - prop.se,
                                ymax = prop + prop.se))+
  labs(title = "Proportion of Home Runs per At-Bat by Age",
       x = "Age",
       y = "Proportion of HR per AB")
```
